\documentclass{article}

\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{colortbl}
\usepackage{amsmath}
\usepackage{float}

\usepackage{listings}
\begin{document}

\title{CM30225 Parallel Computing \\ Assessed Courseork Assignment 1}
\author{}

\maketitle

\section{Approach}
The relaxation problem involves repetadly replacing each cell in a matrix with the
average of its four neighbours until a given precision is reached. Firstly to solve
the probelm using a serial approach each cell is processed row by row. After each
cell has its value replaced the diffrence between the previous and new value is
calculated and checked to see if it is less than the precision value. When every
cell is replaced by a number smaller than the precision the computation stops.\\~\\
In order to store the matrix two double**'s are created one to record the result of
the last iteration and one to store all the new values in, after each iteration the
pointers for the two matrix's are swapped.


\section{Parralisation Technique}
In order to parallelise the problem it was decided to spilt the matrix up into rows
then give each thread a number of these rows. More specifically each thread is
given a starting row, which is applies relaxation too, then adds the number of
threads to the starting row to get the next row to compute on. So if 4 threads
are used on a 14 row matrix, the two end rows are fixed and don't require relaxation,
the rows are split up like so:

\begin{center}
\begin{tabular}{ |c|c|c|c|c|c|c|c|c|c|c|c|c|c|c| }
 \hline
 rowNumber & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 \\
 thread &  & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 &  \\
 \hline
\end{tabular}
\end{center}

This splits the rows evenly between the given number of threads. If the number of
number of threads is more than the size of the matrix then the number of threads
is reset to the size of the matrix.\\~\\
Each thread is allowed to run at its own speed while it computes the relaxation
for its set of rows. The threads are then all syncronised using a barrier before
they can move on to the next iteration. After all rows have been computed the
read and write matrixs must be swapped this is a serial operation so only one
thread performs is while the rest wait. To make sure only one thread performs this
the C function pthread\_barrier\_wait returns the value PTHREAD\_BARRIER\_SERIAL\_THREAD
to one thread, the thread that gets given this is the one which will perform the
serial step.The pseudocode below shows how this is implemented.\\~\\

\begin{lstlisting}
while(precision_not_met)
{
  doRelaxation();

  serial = barrier_wait();

  if (serial == PTHREAD_BARRIER_SERIAL_THREAD)
  {
    doSerialWork();
  }

  barrier_wait();
}
\end{lstlisting}


A quick note is that the program will actual use the number of thread plus the
main thread, it was decided not to change this as the main thread just will be
de-scheduled while it waits for all the other threads to return so won't effect
the investigation significantly.

Finally to check if the precision has been made a global variable is used by all
threads to show whether the program should continue. This global value is set to
0 before each iteration then if any of the threads do not meet the precision they
set it to 1. This means that if by the end of the iteration the value is 1 then
the computation needs to continue.

\newpage
\section{Avoiding Race Conditions}
There are a few places where race conditions are possible:

\begin{enumerate}
\item While thinking about this problem there were two options firstly to read and write
to the same matrix as this would probebly tend towards the same answer. However
as multiple threads could be trying to read and write to the same cell at once the
program could yeild diffrent results each time unless the matrix was protected by a lock, which would add
overhead.
Therefore two matrixs are used are so one can be used to read from
and then one to write to. This avoids threads trying to write and read to the same cell simultaniously,
similtanious reads are possible and not a problem.\\
\item Once a thread has finished it sets a global variable ``cont" to 1 if the
precision was not met. this variable does not need a lock as if one thread resets
it after it doesn't matter as its still 1.
\item After all threads have finished they check if the continue value is 0 and if
so return. As each thread can update this value then all threads must have finished
the iteration before any one can check this value, so a barrier is required before.
Also as the next step is to set the value to 0 and continue the next iteration a barrier
is required after the check as well to make sure that all threads have checked the value
before reseting it.
\item The swapping of the matrix's and resetting the continue back to 0 needs to be
done by a single thread after all threads have finished relaxing. therefore before a
pthread\_barrier\_wait appears before and after this serial code (lines 135 and 142 in the C program).
\end{enumerate}

With this in mind the pseudocode will actually look like this.\\

\begin{lstlisting}
while(true)
{
  doRelaxation();

  barrier_wait();
  if (precision_met)
  {
    return;
  }

  serial = barrier_wait();

  if (serial == PTHREAD_BARRIER_SERIAL_THREAD)
  {
    swapMatrixs();
    precision_met = 0;
  }

  barrier_wait();
}
\end{lstlisting}

\newpage
\section{Correctness Testing}
In order to test the program is correctly computing the answer the problem was split
into three seperate sections which can be tested seperatly. Firstly the relaxation,
in order to test this the first three iterations of a 5 x 5 matrix bounded by all 1's
where hand calculated then compared to the output from my program running with one
thread. For example in the first iteration you would expect 0.5 at the corners (1 + 1 + 0 + 0 / 4)
, 0.25 on the edges (1 + 0 + 0 + 0 / 4) and zero in the middle. As you can see below the program outputs the correct
value.

\begin{center}
\begin{tabular}{ c c c c c }
 1 & 1 & 1 & 1 & 1 \\
 1 & 0.5 & 0.25 & 0.5 & 1 \\
 1 & 0.25 & 0.0 & 0.25 & 1 \\
 1 & 0.5 & 0.25 & 0.5 & 1 \\
 1 & 1 & 1 & 1 & 1 \\
\end{tabular}
\end{center}

This was continued for the next two iterations, after this it was concluded that
the program can correctly calculate each relaxation iteration.\\
The second step to the correctness testing is that the relaxation stops when a given
precision has been reached. To test this a precision is picked, 0.01 for example,
then two 3 x 3 matrix was then chosen with bounding values 0.011 and 0.01 when relaxation
is performed the first matrix should iterate twice and the second only once as the precision
is imediatly met.\\
Below are two tables showing the two runs on the program, firstly with 0.011 bounding
and secondly with 0.01.

\begin{center}
\begin{tabular}{ c c c c }
iteration & diffrence & precision & continue \\
 1 & 0.011 & 0.01 & 1\\
 2 & 0.0 & 0.01 & 0\\
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{ c c c c }
iteration & diffrence & precision & continue \\
 1 & 0.01 & 0.01 & 0\\
\end{tabular}
\end{center}

This was tried with precisions 0.1, 0.01, 0.001 and according bounding values and
then it was concluded that the relaxation correctly stops when the precision is met.\\~\\

The Final thing to test is that all of the above still holds for multiple threads,
to check this various number of threads were run using the same matrix size and the results
were checked for equality. Once this was done for 4x4, 8x8, 10x10 and 20x20 with
2, 4, 8 and 16 threads it was conculded that the program was correctly calculating
the relaxation of a matrix to a given precision with any number of threads.

\section{Scalability Investigation}
\subsection{Speedup}
Speedup is given as the speed taken to preform a computation on one processor
divided by the speed on P processors.\\~\\
\begin{center}
S_p = \frac{Time\ on\ 1\ processor}{Time\ on\ p\ processors}\\~\\
\end{center}
Given this the speedup was calculated for a 10000 by 10000 matrix for all number
of threads between 1 and 16. The results of this are given below.\\~\\
\begin{center}
\pgfplotstabletypeset[
  columns/threads/.style={column name=Threads},
  columns/time/.style={column name=Time (S)},
  columns/speedup/.style={column name=Speedup on P processors},
]{10000x10000}
\end{center}

As you can see from the results the speedup on P processors is very close to the
number of processors until and trails of after 14 processors is reached. This is
shown more easily by Figure \ref{fig:speedup}.\\

\begin{figure}[H]
 \centering
 \begin{tikzpicture}
 \begin{axis}[
     xlabel={Threads},
     ylabel=Speedup,
     ]
   \addplot table [x=threads,y=speedup] {10000x10000};
 \end{axis}
 \end{tikzpicture}
 \caption{10000 x 10000, 0.01 precision}
 \label{fig:speedup}
 \end{figure}\\~\\

 \subsubsection{Amdahl’s Law}
 One of the reasons for the tail off in speedup could be that it is approaching
 Amdahl's Limit. Amdahls' Law states that there is a limit on speedup because
 some part of the program must be done in serial. The serial part of this program
 is the swapping of matrixs and the resetting of the the continue variable.\\
 The theoretical Amdahl Limit was calculated by running the serial part of the code 10,000
 times and getting the average length of time (0.00000472 seconds) the number of
 iterations performed on a 10000 by 10000 matrix with a precision of 0.01 is 37
 this gives the serial portion of the computation a time of 0.00017 Seconds.\\
 In theory this limit would be accesable with an infinate number of processors
 however there are extra overheads which are not taken into account, such as waiting
 on barriers and also createing and destroying the threads. These overheads will also
 increase as more threads are added causing the calculated Amdahl to be impossible to
 get near to with the current program structure.

\subsubsection{Slowdown}

Another thing noticed while doing the scaleability investigation was on smaller
sized arrays, around less than 100 by 100, there was slowdown instead of speedup.
This is shown by Figure \ref{fig:slowdown} which was using a 100 by 100 matrix and 0.01
precision.

\begin{figure}[H]
 \centering
 \begin{tikzpicture}
 \begin{axis}[
     xlabel={Threads},
     ylabel=Time (S),
     ]
   \addplot table [x=threads,y=time] {100x100};
 \end{axis}
 \end{tikzpicture}
 \caption{100 x 100, 0.01 precision}
 \label{fig:slowdown}
\end{figure}\\~\\

As you can see from the graph initially adding more processors does speed the
computation up as you would expect. However after 5 processes are added any more causes
slowdown as the overhead of creating the extra thread outweighs the extra computing
power gained.

\subsubsection{Gustafson’s Law}

Gustafson's law suggests that if the problem size is increased the percentage of time
spent of the sequential part of a program will decrease, therefore giving a better
speed up value. To try and show this in terms of the speedup was calculated for
differing sizes of matrix, these results are shown in Figure \ref{fig:gustafson}.

\begin{figure}[H]
 \centering
 \begin{tikzpicture}
 \begin{axis}[
     xlabel={Array Size},
     ylabel=Speedup,
     ]
   \addplot table [x=arraysize,y=speedup] {gustafson16};
   \addplot table [x=arraysize,y=speedup] {gustafson8};
   \addplot table [x=arraysize,y=speedup] {gustafson4};
   \addlegendentry{16 Cores}
   \addlegendentry{8 Cores}
   \addlegendentry{4 Cores}
 \end{axis}
 \end{tikzpicture}
 \caption{0.01 precision}
 \label{fig:gustafson}
\end{figure}\\~\\

These results show that, as Gustafson suggests, as the problem size is increased
the speedup on P processors tends towards the number of processors.

\subsubsection{Efficiency}

Efficiency, speed up per processor is given by the following equation.\\~\\

\begin{center}
$E\ =\ \frac{time\ on\ 1\ processor}{p\ \times\ time\ on\ P\ processors}$
\end{center}

\begin{figure}[H]
 \centering
 \begin{tikzpicture}
 \begin{axis}[
     xlabel={Threads},
     ylabel=Efficiency,
     ]
   \addplot table [x=cores,y=efficiency] {Efficiency/100x100};
   \addplot table [x=cores,y=efficiency] {Efficiency/1000x1000};
   \addplot table [x=cores,y=efficiency] {Efficiency/10000x10000};
   \addlegendentry{100 x 100}
   \addlegendentry{1000 x 1000}
   \addlegendentry{10000 x 10000}
 \end{axis}
 \end{tikzpicture}
 \caption{0.01 precision}
 \label{fig:efficiency}
\end{figure}\\~\\

As shown in Figure \ref{fig:efficiency} as more threads are added the efficiency
drops, this is expected as the more threads there are the more processor time is
spent waiting for other threads. How fast the efficiency drops is dependant on
the problem size. Using a small problem 100 x 100 the efficiency drops very quickly
as the percentage part of the problem which can be done in parrellel is smaller
so the threads spend a larger portion of time waiting for the serial part to complete, larger
problem sizes drop off slower.

\subsubsection{Karp-Flatt}

The Karp-Flatt metric was calculated for three of the matrix sizes to see how the
diffrence in problem size and number of threads effects the metric. Figure \ref{fig:karpflatt}
shows the results from this investigation.

\begin{figure}[H]
 \centering
 \begin{tikzpicture}
 \begin{axis}[
     xlabel={Threads},
     ylabel=KarpFlatt,
     ]
   \addplot table [x=cores,y=karpflatt] {KarpFlatt/100x100};
   \addplot table [x=cores,y=karpflatt] {KarpFlatt/1000x1000};
   \addplot table [x=cores,y=karpflatt] {KarpFlatt/10000x10000};
   \addlegendentry{100 x 100}
   \addlegendentry{1000 x 1000}
   \addlegendentry{10000 x 10000}
 \end{axis}
 \end{tikzpicture}
 \caption{0.01 precision}
 \label{fig:karpflatt}
\end{figure}\\~\\

As shown by the figure for large problem sizes the Karp-Flatt metric stays roughly
the same as you add threads and is very close to 0. This is expected as the speedup on
a number of processors is very close to the number of processors, this was shown
earlier. However when the problem size is small the Karp-Flatt metric gets larger
as more processors are added, this is because the Karp-Flatt metris is a measure of
the sequential part of the problem and with small problems there is, propotionall,
more sequential part as threads spend longer waiting on each other.

\subsection{Parrellel Overhead}

In order to evaluate the work efficiency of the solution the overhead was calculated
using this formula. This gives the total parrallel overhead in seconds of all threads.
This was calculated for a selection of problem sizes to see how this effected the overhead using diffrenct cores.
To make the data more obvious the Overhead per thread was calculated, the overheaddivided by the number of threads,
this is what is shown in Figure \ref{fig:overhead}
and Figure \ref{fig:overhead2}.

\begin{center}
$Overhead\ =\ p\ \times\ time\ on\ P\ processors\ -\ time\ on\ 1\ processor}$
\end{center}

\begin{figure}[H]
\begin{minipage}{.5\textwidth}
 \centering
 \resizebox{.9\textwidth}{.9\textwidth}{
  \begin{tikzpicture}
  \begin{axis}[
     xlabel={Threads},
     ylabel=Overhead (S),
     ]
   \addplot table [x=cores,y=overhead] {Overhead/5000x5000};
   \addplot table [x=cores,y=overhead] {Overhead/10000x10000};
   \addlegendentry{5000 x 5000}
   \addlegendentry{10000 x 10000}
  \end{axis}
  \end{tikzpicture}
 }
 \caption{0.01 precision}
 \label{fig:overhead}
\end{minipage}
\begin{minipage}{.5\textwidth}
\centering
 \resizebox{.9\textwidth}{.9\textwidth}{
\begin{tikzpicture}
\begin{axis}[
    xlabel={Threads},
    ylabel=Overhead (S),
    ]
  \addplot table [x=cores,y=overhead] {Overhead/500x500};
  \addplot table [x=cores,y=overhead] {Overhead/1000x1000};
  \addlegendentry{500 x 500}
  \addlegendentry{1000 x 1000}
\end{axis}
\end{tikzpicture}
}
\caption{0.01 precision}
\label{fig:overhead2}
\end{minipage}
\end{figure}\\~\\

This data shows that as you add more threads the total overhead will increase. This is
not a supprise as adding more threads requires more time to be spent creating and destroying
the pthreads and also waiting on barriers will have a longer overhead as there are
more threads to wait for.

\subsection{Isoefficiency}

To work out how the program scales the Isoefficiency was calculated. The Isoefficiency
is the relationship between the number of threads and the size of the problem where the
efficiency stays the same. So how much larger does the problem have to be to keep the
efficiency the same for a given increase in the number of threads.\\
In order to analyse the isoefficiency we need an equation for the sequential time
and parrelle time. An equation to give the sequential time is as follows where
n is the size of the matrix, a is some overhead inncured in checking the precision
after each iteration and b is the overhead in setting up the calculation.

\begin{center}
$T_{seq}\ =n^2\ +\ na\ + b$
\end{center}

The parrellel time equation is similar to the sequntial one but the $n^2$ part is done
in parrallel, there is also an overhead c in setting up each thread. So the equation
is as follows. where p is the number of threads.

\begin{center}
$T_{par}\ =\frac{n^2}{p}\ +\ pc\ + na + b$
\end{center}

And the parrellel over is therefore give by this equation.

\begin{center}
$T_{o}\ =p^2c\ +\ (p\ -\ 1)(na\ +\ b)$
\end{center}

So subbing these two equations into $T_{seq}\ =\ C(T_{par})$ and simplifying gives
us.

$n^2\ +\ na\ +\ b\ +\ naC\ +bC = p^2cC\ +\ pnaC\ +\ pb$

The by setting the constants to a set value and trying diffrent values for the
matrix size and calculating p we can see that the relationship between the problem
size and the number of threads is linear. To try and get a better idea of how this
theoretical calculation relates to the actual program the efficency was calculated
for a number of problem sizes, then an efficency was picked and the number of cores
needed for that efficiency on each problem size was found. The table below shows the
findings for some of the efficiencys tested.

\begin{center}
\begin{tabular}{ |c|c|c| }
 \hline
  Problem size & 0.6 precision & 0.7 precision \\
  300 & 7.1 & 5\\
  500 & 11 & 8\\
  700 & 14.2 & 10.5\\
  900 & 16 & 12.1\\
 \hline
\end{tabular}
\end{center}

As shown in Figure \ref{fig:Isoefficiency} the relationship between the number of threads
and the problem size is very close to linear as suggested by the calculations done
above. In this example the 

\begin{figure}[H]
 \centering
 \begin{tikzpicture}
 \begin{axis}[
     xlabel={Problem Size},
     ylabel=Threads,
     ]
   \addplot table [x=size,y=six] {Isoefficiency/100};
   \addplot table [x=size,y=seven] {Isoefficiency/100};
   \addlegendentry{0.6 precision}
   \addlegendentry{0.7 precision}
 \end{axis}
 \end{tikzpicture}
 \caption{}
 \label{fig:Isoefficiency}
\end{figure}\\~\\

\subsubsection{Conclusion}

I conclusion the program


\end{document}
